Conceptualizing open health data platforms for low- and middle income countries

### 2024-05-27 index.qmd
wip - start rewrite table 1

---
title: "Conceptualizing open health data platforms for low- and middle income countries"
format:
  elsevier-html:
    toc: true
    toc-depth: 2
    fig-cap-location: top
    code-fold: true
  elsevier-pdf:
    keep-tex: true
    journal:
      name: JIMR
      cite-style: authoryear
author:
  - name: Daniel Kapitan
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
      - name: Eindhoven University of Technology
        city: Eindhoven
        country: the Netherlands
    orcid: 0000-0001-8979-9194
    email: daniel@kapitan.net
  - name: Femke Heddema
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Julie Fleischer
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Chris Ihure
    affiliations:
      - name: PharmAccess Kenya
        city: Nairobi
        country: Kenya
  - name: Antragama Abbas
    affiliations:
      - name: Delft University of Technology
        city: Delft
        country: the Netherlands
  - name: Steven Wanyee
    affiliations:
      - name: IntelliSOFT
        city: Nairobi
        country: Kenya
  - name: Alessandro Pietrobon 
    affiliations:
      - name: ONA
        city: Nairobi
        country: Kenya
  - name: Ryan Chrichton
    affiliations:
      - name: Jembi Health Systems
        city: Durban
        country: South Africa
  - name: John Grimes
    affiliations:
      - name: Australian e-Health Research Centre
        city: Brisbane
        country: Australia
  - name: Paula van Brakel
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Mark van der Graaf
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Nicole Spieker
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Mark de Reuver
    affiliations:
      - name: Delft University of Technology
        city: Delft
        country: the Netherlands

@@

Case studies on digital platforms in healthcare point to an emerging pattern where the focus shifts from the digital platform with its defining software and hardware components, to the data as the primary object of interest in and of itself [@ozalp2022digital;@alaimo2022organizations]. This development roughly correlates with the commonly used distinction between primary and secondary health data sharing [@cascini2024health]. While people are generally willing to share their health data to receive the best care (primary use), the attitude towards secondary use of health data varies greatly depending on the type and context. The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing, ties in with the proposed research agenda by de Reuver et al. to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness]. Generally, data platforms inherit the characteristics of digital platforms. From an economic perspective, for example, both types exhibit multi-sided markets. At the same time, data platforms differ as their main offerings revolve around data. From an ecosystem perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity.

Particularly relevant in the context of health data platforms (HDPs), is the conceptualization of openness. The shift in perspective from digital platforms to data platforms coincides with the paradox of open [@keller2021paradox]. Originally, openness of digital platforms focused on open source, open standards and copyrights, which by has been superseded by “... conflicts about privacy, economic value extraction, the emergence of artificial intelligence, and the destabilizing effects of dominant platforms on (democratic) societies. Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox]. These conflicts are particularly salient in the healthcare domain, as is evidenced by the polemic surrounding data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond]. Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].

@@

This paper is motivated by the conflation of a number of developments relevant to the design and implementation of open, solidarity-based HDPs in LMICs. First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one]. These countries have, as a matter of course, extended the framework to include “analytics services” as an additional domain. The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc. which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair]. In doing so, however, we have implicitly moved from conceptualizing digital health platforms (the original OpenHIE specification) to health data platforms. This is problematic because the notion of openness, which is assumed to be essential in establishing solidarity-based approaches to data sharing, is inherently different for a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open digital platform. Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods]. Typically, these open source components are intended to support deployment in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^1]  If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for data sharing and reuse. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.  The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the four types of data sharing as defined by de Reuver at al. (@tbl-types-data-sharing).

|   | __Type of data sharing__ | __Technology enablers to create openness of health data platforms__ |
|:-:|:-------|:-------------|
| __1__ | Data at the most granular (patient) level, which is persisted and used to provide a longitudinal record. | __Bulk FHIR API__: has by now been incorporated in all major FHIR implementations the FHIR standard can be readily used to support data sharing to patient-level data across a patient population [@mandl2020push;@jones2021landscape]. |
| __2__ | Aggregated data, for example statistics for policy evaluation and benchmarking |__SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows. Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items. Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
| __3__ | Data analytics modules, that provide access to work and access the data. | __Federated learning (FL)__ [@rieke2020future] and __privacy- enhancing technologies (PETs)__ [@scheibner2021revolutionizing;@jordan2022selecting]: new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself. Models can be trained on combined datasets and made available as open source artifacts for decision support. Data analysts can use FL and PETs to work with the data in a collaborative, decentralized fashion. |
| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | __ONNX__ [@onnx]: an open format built to represent machine learning models. |

@@

 1. the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/))
 2. the ONA Canopy platform [https://ona.io/home/services/canopy/](https://ona.io/home/services/canopy/)
 3. the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).

@@

While the lakehouse architecture separates the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further. A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A, IR) and Apache Arrow for data connectivity and data memory layout (standards B and C, respectively). The composable data stack can be implemented with current open source technologies (@fig-cds-examples). As an example, the Ibis user interface is currently sufficiently mature to offer a standardized dataframe interface to 19 different execution engines.

@@

Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods. To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2]. Besides the core functional components of the OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and integration of generic components such as Identity and Access Managment (IAM) and a reverse proxy gateway. In the following, we will evaluatie three of such configurations, with the aim to conceptualize and evaluate the proposed Data and Analytics Services domain of of the OpenHIE architecture.

>>>>

---
title: "Conceptualizing open health data platforms for low- and middle income countries"
format:
  elsevier-html:
    toc: true
    toc-depth: 2
    fig-cap-location: top
    code-fold: true
  elsevier-pdf:
    keep-tex: true
    journal:
      name: JIMR
      cite-style: authoryear
author:
  - name: Daniel Kapitan
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
      - name: Eindhoven University of Technology
        city: Eindhoven
        country: the Netherlands
    orcid: 0000-0001-8979-9194
    email: daniel@kapitan.net
  - name: Femke Heddema
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Julie Fleischer
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Chris Ihure
    affiliations:
      - name: PharmAccess Kenya
        city: Nairobi
        country: Kenya
  - name: Steven Wanyee
    affiliations:
      - name: IntelliSOFT
        city: Nairobi
        country: Kenya
  - name: Alessandro Pietrobon 
    affiliations:
      - name: ONA
        city: Nairobi
        country: Kenya
  - name: Ryan Chrichton
    affiliations:
      - name: Jembi Health Systems
        city: Durban
        country: South Africa
  - name: John Grimes
    affiliations:
      - name: Australian e-Health Research Centre
        city: Brisbane
        country: Australia
  - name: Paula van Brakel
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands

  - name: Mark van der Graaf
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Nicole Spieker
    affiliations:
      - name: PharmAccess Foundation
        city: Amsterdam
        country: the Netherlands
  - name: Antragama Abbas
    affiliations:
      - name: Delft University of Technology
        city: Delft
        country: the Netherlands
  - name: Mark de Reuver
    affiliations:
      - name: Delft University of Technology
        city: Delft
        country: the Netherlands

@@

Case studies on digital platforms in healthcare point to an emerging pattern where the focus shifts from the digital platform with its defining software and hardware components, to the data as the primary object of interest in and of itself [@ozalp2022digital;@alaimo2022organizations]. This observation ties into with the proposed research agenda by de Reuver et al. to consider data platforms as a phenomenon distinct from digital platforms [@dereuver2022openness]. Generally, data platforms inherit the characteristics of digital platforms. From an economic perspective, for example, both types exhibit multi-sided markets. At the same time, data platforms differ as their main offerings revolve around data. From an ecosystem perspective, data platforms have more moderate network effects and are more susceptible to fragmentation and heterogeneity.

Particularly relevant in the context of health data platforms (HDPs), is the conceptualization of openness. The shift in perspective from digital platforms to data platforms coincides with the paradox of open [@keller2021paradox]. Originally, openness of digital platforms focused on open source, open standards and copyrights, which by has been superseded by “... conflicts about privacy, economic value extraction, the emergence of artificial intelligence, and the destabilizing effects of dominant platforms on (democratic) societies. Instead of access to information, the control of personal data has emerged in the age of platforms as the critical contention.” [@keller2021paradox]. These conflicts are particularly salient in the healthcare domain, where people are generally willing to share their health data to receive the best care (primary use), while the attitude towards secondary use of health data varies greatly depending on the type and context [@cascini2024health]. The shift in perspective from digital platforms supporting primary data sharing toward data platforms supporting secondary data sharing is one of the key issues surrounding the polemic of data spaces [@otto2022designing] and data solidarity [@kickbusch2021lancet;@prainsack2022data;@prainsack2023beyond]. Openness is particularly relevant if we are to realize a solidarity-based approach to health data sharing that i) gives people a greater control over their data as active decision makers; ii) ensures that the value of data is harnessed for public good; and iii) moves society towards equity and justice by counteracting dynamics of data extraction [@prainsack2022data].

@@

This paper is motivated by the conflation of a number of developments relevant to the design and implementation of open, solidarity-based HDPs in LMICs. First, the OpenHIE framework [@openhie] has been adopted by many sub-Saharan African countries [@mamuye2022health] as the architectural blueprint for implementing nation-wide health information exchanges (HIE), including Nigeria [@dalhatu2023paper], Kenya [@thaiya2021adoption] and Tanzania [@nsaghurwe2021one]. These countries have, as a matter of course, extended the framework to include “data & analytics services” as an additional domain. The rationale for this addition is to facilitate secondary reuse of health data for academic research, real-world evidence studies etc. which can be framed within the context of ongoing efforts towards Findable, Accessible, Interoperable and Reusable (FAIR) sharing of health data [@guillot2023fair]. In doing so, however, we have implicitly moved from conceptualizing digital health platforms for primary data sharing (the original OpenHIE specification) to health data platforms for secondary data sharing. This is problematic because the notion of openness, which is assumed to be essential in establishing solidarity-based approaches to data sharing, is inherently different for a data platform compared to a digital platform.

Conceptually, the OpenHIE framework constitutes a framework for an open digital platform. Openness for digital platforms refers to i) the use of open boundary resources, that is, specifications for the various healthcare specific workflows and information standards such as FHIR; and ii) the use of open source components that are available as digital public goods [@digitalpublicgoods]. If we are to use the OpenHIE framework as an open data platform, we need to extend the standards, technologies and architecture to include functionality for secondary data sharing and reuse. The lack of detailed specifications and consensus of this addition to OpenHIE currently stands in the way of development projects that aim to establish HDPs in LMICs.  The purpose of this paper, therefore, is to specificy how new standards and technologies can be integrated into the OpenHIE architecture such that an open HDP can be realized that supports the four types of data sharing as defined by de Reuver at al. (@tbl-types-data-sharing).

|   | __Type of secondary data sharing__ | __Technology enablers to create openness of health data platforms__ |
|:-:|:-------|:-------------|
| __1__ | Data at the most granular (patient) level, which is persisted and used to provide a longitudinal record. | __Bulk FHIR API__: has by now been incorporated in all major FHIR implementations the FHIR standard can be readily used to support data sharing to patient-level data across a patient population [@mandl2020push;@jones2021landscape]. |
| __2__ | Aggregated data, for example statistics for policy evaluation and benchmarking |__SQL-on-FHIR specification__ [@sql-on-fhir]: provides a standardised approach to make FHIR work well with familiar and efficient SQL engines that are most commonly used in analytical workflows. Builds on FHIRPath [@hl72020fhirpath] expressions in a logical structure to specify things like column names and unnested items. Implementations of this approach are available or forthcoming, including open source implementations such as Pathling [@grimes2022pathling] and commercial offerings like Aidbox [@aidbox].|
| __3__ | Data analytics modules, that provide access to work and access the data. | __Federated learning (FL)__ [@rieke2020future] and __privacy- enhancing technologies (PETs)__ [@scheibner2021revolutionizing;@jordan2022selecting]: new paradigms that address the problem of data governance and privacy by training algorithms collaboratively without exchanging the data itself. Models can be trained on combined datasets and made available as open source artifacts for decision support. Data analysts can use FL and PETs to work with the data in a collaborative, decentralized fashion. |
| __4__ | Trained models that have been derived from the data and can be used stand-alone for decision support. | __ONNX__ [@onnx]: an open format built to represent machine learning models. |

@@

 1. the OpenHIM platform ([https://jembi.gitbook.io/openhim-platform/](https://jembi.gitbook.io/openhim-platform/))
 2. the OnaData platform [https://ona.io/home/products/ona-data/features/](https://ona.io/home/products/ona-data/features/)
 3. the work conducted at PharmAccess Foundation as part of the MomCare programme ([https://health-data-commons.pharmaccess.org](https://health-data-commons.pharmaccess.org)).

@@

While the lakehouse architecture separates the concerns of compute and storage, the composable data stack takes the separation of concerns is taken one step further. A composable data system (@fig-composable-data-stack), not only separates the storage (layer 3) and execution (layer 2), but also separates the user interface (layer 1) from the execution engine by introducing standards including Substrait for Intermediate Representation (standard A, IR) and Apache Arrow for data connectivity and data memory layout (standards B and C, respectively). The first generation of open source components are already available. For example, the Ibis user interface is currently sufficiently mature to offer a standardized dataframe interface to 19 different execution engines [@ibisproject].

@@

Today, many components of the OpenHIE specification are now available as a digital public goods, as listed in @tbl-digital-public-goods. Typically, these open source components are intended to support deployments in small countries (population up to 10 million) or large NGOs out of the box, and should provide a stepping stone for customized deployments in medium-sized countries (population around 40 million).[^1] To further ease the development, configuration and deployment of health information exchanges, the concept of 'Instant OpenHIE' has been championed to (i) allow implementers to engage with a preconfigured health information exchange solution and running tools (based on the architecture) and test their applicability and functionality with a real health context problem; and (ii) have a packaged reference version of the OpenHIE architecture that is comprised of a set of reference technologies and other appropriate tools that form the building blocks of the health information exchange that can be configured and extended to support particular use cases [@InstantOpenHIEv2]. Besides the core functional components of the OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and integration of generic components such as Identity and Access Managment (IAM) and a reverse proxy gateway. In the following, we will evaluatie three of such configurations, with the aim to conceptualize and evaluate the proposed Data and Analytics Services domain of of the OpenHIE architecture.
